### 2024-01-10 globals/metadata.tex
presque un papier

\title{A corrected version of Karhumäki’s theorem
       on commutative polynomial $\Nat$-rational series}
\author{
    Lopez Aliaume
}
\date{\today}
\bibliography{globals/papers}

@@

\newcommand{\makeabstract}{
    \begin{abstract}
        Deciding whether a $\Rel$-rational series
        can be computed as a $\Nat$-rational series
        is an open problem that was solved by Karhumäki in the
        specific case of $\Rel$-rational series that are commutative
        and compute a polynomial.
        We propose a short counter-example to this
        previous result on commutative $\Nat$-rational series,
        and prove a corrected version of the theorem, that only
        requires commutativity and a polynomial growth rate.
        We then discuss how the syntactic criterion could
        pave the way towards a decision procedure when the commutativity
        of the $\Rel$-rational series
        is not assumed.
    \end{abstract}
}

>>>>

\title{
    Star-free commutative $\Nat$-polyregular functions:
    correcting a result of Karhumäki.
}
\author{
    Lopez Aliaume
}
\date{\today}
\bibliography{globals/papers}

@@

\newcommand{\makeabstract}{
    \begin{abstract}
        Deciding whether a $\Rel$-rational series
        can be computed as a $\Nat$-rational series
        is an open problem that was solved by Karhumäki in the
        specific case of $\Rel$-rational series that are commutative
        and compute a polynomial.
        We propose a short counter-example to this
        previous result on commutative $\Nat$-rational series,
        and prove a corrected version of the theorem,
        working of tdd
        We then discuss how the syntactic criterion could
        pave the way towards a decision procedure when the commutativity
        of the $\Rel$-rational series
        is not assumed.

        \begin{itemize}
            \item Deciding $\Nat$ inside $\Rel$.
            \item Counter example.
            \item Characterization of $\Nat$-polyregular.
            \item Characterization of star free $\Nat$-polyregular.
        \end{itemize}
    \end{abstract}
}

### 2024-01-10 pnrs.tex
presque un papier

%! TeX program = xelatex
\documentclass[a4paper,11pt]{article}

@@

@@

@@

\begin{definition}[Commutative rational series]
    A $\Rel$-rational series
    $f \colon \Sigma \to \Rel$ is \intro{commutative} whenever $f(uv) = f(vu)$ for
    all $u,v \in \Sigma^*$.
\end{definition}

@@

\begin{fact}[see e.g. ...]
    \label{polynomial-rational-polyreg:fact}
    Let $\Sigma$ be a finite alphabet, and $f \colon \Sigma^* \to \Nat$
    be a function. The following are equivalent:
    \begin{enumerate}
        \item $f$ can be computed as a \k{$\Nat$-rational series}
            of \kl{polynomial growth}.
        \item $f$ is a \kl{$\Nat$-polyregular function}.
    \end{enumerate}
\end{fact}

@@

\begin{remark}
    In the proof, we actually conclude that
    the \kl{$\Nat$-rational polynomials}
    are exactly those that can be computed 
    by \kl{commutative}
    \emph{star-free} $\Nat$-polyregular functions.
\end{remark}

@@

@@

    Furthermore, the polynomials are computable from $f$.
\end{lemma}
\begin{proof}
    equivalent partitions of the word, but commutativity?.q
\end{proof}

@@

\begin{definition}[Ultimately polynomial]
\end{definition}

\begin{fact}
    Any $\Nat$-rational polynomial is a star-free polyregular function.
\end{fact}

\begin{fact}
    One can do anything on small values.
\end{fact}

Let us now conclude with a nice characterisation.
\begin{theorem}
    If
    $f$ is star-free $\Rel$-polyregular
    and $f$ is commutative $\Nat$-polyregular,
    then $f$ is star-free $\Nat$-polyregular.
\end{theorem}
\begin{proof}
    Idea. 
    Remark that $f$ is ultimately polynomial, and because $f$ 
    is commutative, this proves that $f$, after some rank,
    computes a polynomial. But this proves that $f$ is star-free
    $\Nat$-polyregular.
    (maybe we need to do an induction on the degree).
\end{proof}

\begin{corollary}
    One can decide if a commutative $\Nat$-polyregular function 
    is star-free.
\end{corollary}

>>>>

%! TeX program = xelatex
%! lang = en-GB
\documentclass[a4paper,11pt]{article}

@@

\section{TODOs}

\begin{itemize}
    \item Use $\Nat$-polyregular functions all the time, and just 
        in the introduction talk about $\Nat$-rational series.
    \item Prove that $\Nat$-rational series of polynomial 
        growth are $\Nat$-polyregular functions, maybe using
        the results about ambiguity in automata?
        Theorem 5.22
    \item Define commutativity by a map 
        $\commute \colon \Sigma^* \to \Nat^{\Sigma}$.
    \item Define "commutative functions" and its symbol.
    \item Define "computed by" (like, recognised by) for commutative 
        functions.
    \item Change \cref{corrected-version:thm} to use directly
        star-free!
    \item Clear up the computability problems.
    \item Polynomial + Npolyregular => star free?
    \item Polynomial + N rational => N polyregular?
    \item cite the paper of Quantitative Monadic Second-Order Logic
        about why it is interesting class.
\end{itemize}

@@

\begin{fact}[{\cite[Exercise 1.2 p 169]{berstel2011noncommutative}}]
    \label{polynomial-rational-polyreg:fact}
    Let $\Sigma$ be a finite alphabet, and $f \colon \Sigma^* \to \Nat$
    be a function. The following are equivalent:
    \begin{enumerate}
        \item $f$ can be computed as a \k{$\Nat$-rational series}
            of \kl{polynomial growth}.
        \item $f$ is a \kl{$\Nat$-polyregular function}.
    \end{enumerate}
\end{fact}

\begin{definition}[Letter Counting Function]
    \label{commutative-map:def}
    Let $\Sigma$ be a finite alphabet. We define
    $\commute \colon \Sigma^* \to \Nat^\Sigma$ for the
    map that counts occurrences of every letter in the input word, namely:
    \begin{equation*}
        \commute[w] \defined \seqof{\card[a]{w}}{a \in \Sigma} \quad .
    \end{equation*}
\end{definition}

\begin{definition}[Commutative Function]
    \label{commutative:def}
    Let $X$ be a set.
    A function $f \colon \Sigma^* \to X$ is
    \intro{commutative} whenever
    for all $u,v \in \Sigma^*$, $f(uv) = f(vu)$.

    Equivalently, it is \reintro{commutative} whenever
    there exists a map $g \colon \Nat^\Sigma \to X$
    such that $g \circ \commute = f$.
\end{definition}

\begin{definition}[Represented Function]
    A function $g \colon \Nat^n \to \Rel$
    is \intro{represented} by a function $f \colon \Sigma^* \to \Rel$
    if there exists a map $\mu \colon \set{1, \dots, n} \to \Sigma$
    such that
    $g \circ (v \mapsto \seqof{v_{\mu(k)}}{1 \leq k \leq n}) \circ \commute = f$.
\end{definition}

We need, $\ZRat$, $\NRat$, $\ZPoly$, $\NPoly$, $\ZSF$, and $\NSF$.

@@

@@

@@

\begin{theorem}
    \label{nsf-polynomials:thm}
    In the proof, we actually conclude that
    the \kl{$\Nat$-rational polynomials}
    are exactly those that can be computed 
    by \kl{commutative}
    \emph{star-free} $\Nat$-polyregular functions.
\end{theorem}

@@

\begin{lemma}
    f is commutative $\Nat$ polyregular, then when iterated we obtain
    polynomials in $\CorrectPoly$.
\end{lemma}
\begin{proof}
    by induction on the number of letters 
\end{proof}

@@

    Furthermore, the polynomials are computable from $f$.
\end{lemma}
\begin{proof}
    equivalent partitions of the word, but commutativity?.q
    Lemma 5.37 phd gaëtan.
\end{proof}

@@

Let us now prove that the above characterizations of
commutative
$\Nat$-polyregular functions can be combined with the recent
advances in the study of $\Rel$-polyregular functions 
\cite{LOPEZ23b}
to prove the following theorem.

\begin{theorem}
    \label{zsf-npoly-nsf:thm}
    Let $\Sigma$ be a finite alphabet, and $f \colon \Sigma^* \to \Nat$
    be both a \kl{star-free $\Rel$-polyregular function}
    and a \kl{commutative} \kl{$\Nat$-polyregular function}.
    Then, $f$ is a \kl{commutative} \kl{star-free $\Nat$-polyregular function}.
    That is, the following equality holds:
    \begin{equation*}
        \ZCommut \cap \ZSF \cap \NPoly
        = \ZCommut \cap \NSF
        \quad .
    \end{equation*}
\end{theorem}

Note that this proves in the commutative case a conjecture that was 
formulated by Thesis Gaetan, and that is non trivial, because

\begin{conjecture}
    \label{nsf-zsf:conj}
    In the general case, the following equation holds:
    \begin{equation*}
        \ZSF \cap \NPoly
        = 
        \NSF
        \quad .
    \end{equation*}
\end{conjecture}

As a corollary of \cref{zsf-npoly-nsf:thm} and the effective decision
procedures of $\ZSF$ obtained in \cite{LOPEZ23b}, we conclude that $\NSF$ is
decidable inside $\NPoly$ under the assumption of commutativity.

\begin{corollary}
    One can decide if a \kl{commutative} \kl{ $\Rel$-polyregular function}
    can be realized by a \kl{star-free $\Nat$-polyregular function},
    in which case the conversion is effective.
\end{corollary}

The key ingredient of this section is the use of a semantic characterization of
\kl{star-free $\Rel$-polyregular functions} among \kl{$\Rel$-rational series}
that generalizes the aperiodicity of languages to functions by the means of
polynomial behaviors.

\begin{definition}[Ultimately polynomial]
    \label{ultimately-polynomial:def}
    Let $\Sigma$ be a finite alphabet. 
    A function $f \colon \Sigma^* \to \Rel$
    is \intro{ultimately polynomial}
    whenever there exists $N_0 \in \Nat$ such that
    for all $n \in \Nat$
    and for all words $\alpha_0, w_1, \alpha_1, \cdots, \alpha_{n-1}, w_n, \alpha_n
    \in \Sigma^*$, there exists a polynomial $P \in \Rel[X_1, \dots, X_n]$
    such that
    \begin{equation*}
        f\left(
            \alpha_0 \prod_{i = 1}^{n} w_i^{X_i} \alpha_i
        \right)
        = 
        P(X_1, \dots, X_n)
        \quad 
        \forall X_1, \dots, X_n \geq N_0
        \quad .
    \end{equation*}
\end{definition}

Let us recall from \cite{LOPEZ23b} that this definition correctly
generalizes the indicator functions of regular languages.

\begin{example}
    A language $L$ is aperiodic if and only if 
    the indicator function is ultimately polynomial.
\end{example}
\begin{proof}
    TODO.
\end{proof}

\begin{proof}[Proof of \cref{zsf-npoly-nsf:thm}]
    We proceed by induction on the size of the alphabet $\card{\Sigma}$.

    \textbf{Base Case.} When $\Sigma$ is empty, $\Sigma^*$ contains
    only the empty word $\varepsilon$, and because $f$ is a
    \kl{$\Nat$-polyregular function}, $f(\varepsilon) \in \Nat$.
    Note that $f$ is therefore a constant function, equal to a natural
    number, hence belongs to $\NSF$.

    \textbf{Induction.}
    Because $f$ is \kl{ultimately polynomial},
    there exists $N_0 \in \Nat$, 
    and $P \in \Rel[\seqof{X_a}{a \in \Sigma}]$ 
    such that
    \begin{equation*}
        f\left( \prod_{a \in \Sigma} a^{X_a} \right)
        = 
        P(\seqof{X_a}{a \in \Sigma})
        \quad
        \text{ when }
        \forall a \in \Sigma, X_a \geq N_0
        \quad .
    \end{equation*}
    Furthermore, leveraging \cref{decompose-polynomial:lem},
    we obtain $\omega \in \Nat$
    and a family of polynomials 
    $\seqof{P_{\seqof{n_a}{a \in \Sigma}}}{0 \leq n_a < \omega}$
    such that.
    Because each of those polynomials coincide with $P$ over an infinite
    subgrid of $\Nat$,
    we conclude that all polynomials are in fact equal to a single one.
    we conclude that the polynomials 
    $P_{\seqof{n_a}{a \in \Sigma}}$ are all equal.

    Now, we conclude in particular that the function 
    computing $P$ is \kl{$\Nat$-polyregular}, because $f$ is,
    hence is computable by a \kl{star-free $\Nat$-polyregular function},
    because of \cref{nsf-polynomials:thm}.
    When fixing values of some indeterminate below $N_0$,
    we obtain a new function $g$, that remains commutative,
    remains ultimately polynomial, and remains \kl{$\Nat$-polyregular}.
    By induction hypothesis, this function $g$ is also 
    \kl{star-free $\Nat$-polyregular}, and we conclude
    that 
    $f = P + \sum g$ which remains star free.
\end{proof}
